Notes on Control Plane Inference

### 2021-03-29 inference.tex
Add subgrammars

% sets
\newcommand{\Value}{\mathsf{Value}}
\newcommand{\BVExpr}{\mathsf{BVExpr}}
\newcommand{\BExpr}{\mathsf{BExpr}}
\newcommand{\Cmd}{\mathsf{Cmd}}
\newcommand{\Pkt}{\mathsf{Pkt}}
\newcommand{\Model}{\mathsf{Model}}
\newcommand{\Hdr}{\mathsf{Hdr}}
\newcommand{\Field}{\mathsf{Field}}
\newcommand{\Table}{\mathsf{Table}}
\newcommand{\Row}{\mathsf{Row}}
\newcommand{\List}{\mathsf{List}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\SymbRow}{\mathsf{SymbRow}}

@@

% math
\newcommand{\argmin}{\mathop{\textrm{argmin}}}

@@

% Theorems
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

@@

\begin{figure}[htp]
  \[\begin{array}{lcll}
    \multicolumn{3}{l}{\Value} \\
    v & ::= & [n]_w  & \textit{Bitvector Literal}\\
    \multicolumn{3}{l}{\SymbRow} \\
    \rho & ::= & \left\{
    \begin{array}{l}
      \matches = \List[x];\\
      \action = x;\\
      \data = \List[x]
    \end{array}\right\} & \textit{Symbolic Row}\\
    \multicolumn{3}{l}{\BVExpr} \\
    e & ::= & v & \textit{Value}\\
      & \mid & h.f & \textit{Header Access} \\
      & \mid & \rho.\matches[i] & \textit{Symbolic Match} \\
      & \mid & \rho.\action & \textit{Symbolic Action}\\
      & \mid & \rho.\data[i] & \textit{Symbolic Action Data}\\
      & \mid & x & \textit{Variable}\\
      & \mid & e \binop e & \textit{Arbitrary Binary Operation}\\
    & \mid & \unop e & \textit{Arbitrary Unary Operation}\\
    \multicolumn{3}{l}{\Table} \\
    t & ::= & \left\{\begin{array}{l}
    t.\id = v; \\
    t.\keys = \List[h.f]; \\
    t.\actions = \List[\lambda \vec x. c]\\
    \end{array}\right\} & \textit{Table}\\
    \multicolumn{3}{l}{\BExpr} \\
    b & ::= & \FALSE  & \textit{Falsity}\\
      & \mid & b \Rightarrow b & \textit{Implication} \\
      & \mid & e = e & \textit{Bitvector Equality}\\
    & \mid & \forall \rho \in t.~b & \textit{Row Universality}\\
    \multicolumn{3}{l}{\Cmd} \\
    c & ::=  & h.f := e & \textit{Assignment}\\
      & \mid & \assert b & \textit{Assertion}\\
      & \mid & \assume b & \textit{Assumption}\\
      & \mid & t.\apply() & \textit{Table Application}\\
      & \mid & c;c & \textit{Sequential Composition} \\
    & \mid & c \choice c & \textit{Nondeterministic Choice}\\
    & \mid & \choose \rho t c & \textit{Selection} \\
    \multicolumn{3}{l}{
      \begin{array}{ll}
        x \in \Var & h \in \Hdr\\
        n,w,i \in \mathbb{N} & f \in \Field
    \end{array}} & \textit{Sets }\\
  \end{array}
\]
\caption{Grammar of Dataplane Programs}
\label{fig:grammar}
\end{figure}

@@

For actions, we restrict the language to be generated by only non-symbolic
assignment, sequence, and $\SKIP$.

Define $\tables(c) \subset \Table$ to be the set of all tables hat occur in $c$.

@@

\begin{figure}
  \[\begin{array}{r>{\triangleq}cl}
  (h.f := e)[\sigma] && h.f := e   \\
  (\assume b)[\sigma] && \assume b \\
  (\assert b)[\sigma] && \assert b \\
  (c_1;c_2)[\sigma] && c_1[\sigma];c_2[\sigma] \\
  (c_1 \choice c_2)[\sigma] && c_1[\sigma] \choice c_2[\sigma] \\
  (\choose \rho t c)[\sigma] && \textrm{\textcolor{red}{unnecessary, but similar to t.apply}}\\
  (t.\apply())[\sigma] &&
  \begin{array}[t]{l}
    \{\\
    \quad \assume {r_1.\matches = t.\keys};\\
    \quad t.\action[r_1.\action](r_1.\data)\\
    \}~\choice\\
    \multicolumn{1}{c}{\vdots}\\
    \choice~\{\\
    \quad \assume {r_n.\matches = t.\keys};\\
    \quad t.\action[r_n.\action](r_n.\data)\\
    \}\\
    \textrm{where}~[r_1;\ldots;r_n] = \sigma(t.\id)
  \end{array}
  \end{array}
  \]
  \caption{Table Elimination}
  \label{fig:table-elim}
\end{figure}

@@

\begin{figure}[htp]
  \[\begin{array}{r >{\triangleq}cl}
  (h.f := e)[?] && h.f := e \\
  (\assume b)[?] && \assume b \\
  (\assert b)[?] && \assert b\\
  (c_1;c_2)[?] && c_1[?];c_2[?]\\
  (c_1 \choice c_2)[?] && c_1[?] \choice c_2[?]\\
  (\choose \rho t c)[?] &&
  \textrm{\textcolor{red}{unnecessary}} \\
  (t.\apply())[?] &&
  \begin{array}[t]{l} \displaystyle
    \{ \\
    \quad \choose \rho t \;\{ \\
    \qquad \assume \rho.\matches = t.\keys; \\
    \qquad \displaystyle \bigchoice_{i=0}^{|t.\action| -1}
    \left\{\begin{array}{l}
    \assume \rho.\action = i;\\
    t.\action[i](\rho.\data)
    \end{array}\right\}\\
    \quad \} \\
    \} \choice \{ \\
    \quad \assume \forall \rho \in t. \rho.\matches \neq t.\keys; \\
    \quad \SKIP \\
    \quad \}\\
    \}
  \end{array}
  \end{array}
  \]
  \caption{Symbolic Table instrumentation}
  \label{fig:table-instrument}
\end{figure}

\begin{lemma}[Models Invariant over Actions]
  \label{lem:models-invariant-over-actions}
  For an action command $c$, $\mu,\mu'
  \in \Model$, $\sigma \in \Value \to
  \Row^*$, $\pkt \in \Pkt$, there exists some $\pkt'$
  \[\denote{c}^\sigma\,\mu\,\pkt = \{\mu,\pkt'\}\]
  \end{lemma}

\begin{proof}
  Proceed by induction over $c$.
  \begin{enumerate}[align=left]
  \item[($c = \SKIP$)] Immediate.
  \item[$(c = h.f = e)$] Immediate.
  \item[$(c = c_1;c_2)$]
    By IH($c_1$) there exists some $\pkt''$ such that
    $\denote{c_1}^\sigma\,\mu\,\pkt = \{(\mu,\pkt'')\}$.
    By IH($c_2$) there exists some $\pkt'$ such that
    $\denote{c_2}^\sigma,\mu\,pkt'' = \{(\mu, \pkt')\}$.
    The result follows by definition of $\denote{-}$.
    \end{enumerate}
\end{proof}

\begin{theorem}[Instrumentation Expressivity]
  For every $c \in \Cmd$, well-formed state $\sigma : \Value \to \Row^*$, model
  $\mu \in \Model$, and packet $\pkt \in \Pkt$, $\denote{c}^\sigma~\mu~\pkt =
  \denote{c[?]}^\sigma~\mu~\pkt$.
\end{theorem}

@@

      By Lemma~\ref{lem:models-invariant-over-actions}, there's some $\pkt'$ such that
      \[\{\mu\{\rho \mapsto r_i\}, \pkt'\} =
      \denote{c[r_i.\data/\vec x]}^\sigma~(\mu\{\rho \mapsto r_i\})~\pkt
      \]

      which means that the result of the first branch is
      \[\{\mu, \pkt'\} =
      \denote{c[r_i.\data/\vec x]}^\sigma~(\mu\{\rho \mapsto r_i\})~\pkt
      \]

      Since $\rho$ doesn't occur in $c$, then also
      \[\{\mu, \pkt'\} = \denote{c[r_i.\data/\vec x]}^\sigma~\mu~\pkt \]

@@

    \item[\textsc{Miss}]
      \todo[inline]{Should be easy}

@@

>>>>

% sets
\newcommand{\Value}{\mathsf{Value}}
\newcommand{\BVExpr}{\mathsf{BVExpr}}
\newcommand{\BExpr}{\mathsf{BExpr}}
\newcommand{\Cmd}{\mathsf{Cmd}}
\newcommand{\Action}{\mathsf{Action}}
\newcommand{\Instr}{\mathsf{Instr}}
\newcommand{\Prog}{\mathsf{Prog}}
\newcommand{\GCL}{\mathsf{GCL}}
\newcommand{\Pkt}{\mathsf{Pkt}}
\newcommand{\Model}{\mathsf{Model}}
\newcommand{\Hdr}{\mathsf{Hdr}}
\newcommand{\Field}{\mathsf{Field}}
\newcommand{\Table}{\mathsf{Table}}
\newcommand{\Row}{\mathsf{Row}}
\newcommand{\List}{\mathsf{List}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\SymbRow}{\mathsf{SymbRow}}

@@

% math
\newcommand{\argmin}{\mathop{\textrm{argmin}}}
\newcommand{\WP}{\textit{wp}}

@@

% Theorems
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{conjecture}{Conjecture}

@@

\begin{figure}[htp]
  \[\begin{array}{lclll}
    \multicolumn{3}{l}{\Value} \\
    v & ::= & [n]_w  & \textit{Bitvector Literal} & (a,g,p,\iota) \\
    \multicolumn{3}{l}{\SymbRow} \\
    \rho & ::= & \left\{
    \begin{array}{l}
      \matches = \List[x];\\
      \action = x;\\
      \data = \List[x]
    \end{array}\right\} & \textit{Symbolic Row} & (\iota)\\
    \multicolumn{3}{l}{\BVExpr} \\
    e & ::= & v & \textit{Value} & (a,g,p,\iota)\\
      & \mid & h.f & \textit{Header Access} & (a,g,p,\iota)\\
      & \mid & \rho.\matches[i] & \textit{Symbolic Match} & (\iota)\\
      & \mid & \rho.\action & \textit{Symbolic Action} & (\iota) \\
      & \mid & \rho.\data[i] & \textit{Symbolic Action Data} & (\iota) \\
      & \mid & x & \textit{Action Data Variable} & (a^*) \\
      & \mid & e \binop e & \textit{Arbitrary Binary Operation} & (a,g,p,\iota) \\
    & \mid & \unop e & \textit{Arbitrary Unary Operation} & (a,g,p,\iota) \\
    \multicolumn{3}{l}{\Table} \\
    t & ::= & \left\{\begin{array}{l}
    t.\id = v; \\
    t.\keys = \List[h.f]; \\
    t.\actions = \List[\lambda \vec x.\,a]\\
    \end{array}\right\} & \textit{Table} & (p)\\
    \multicolumn{3}{l}{\BExpr} \\
    b & ::= & \FALSE  & \textit{Absurdity}& (a,g,p,\iota)\\
      & \mid & b \Rightarrow b & \textit{Implication} & (a,g,p,\iota)\\
      & \mid & e = e & \textit{Bitvector Equality} & (a,g,p,\iota)\\
    & \mid & \forall \rho \in t.~b & \textit{Row Universality} & (\iota)\\
    \multicolumn{4}{l}{c \in \Cmd, a \in \Action, g \in \GCL, p \in \Prog, \iota \in \Instr } \\
    c & ::=  & h.f := e & \textit{Assignment} & (g,p,\iota) \\
      & \mid & \assert b & \textit{Assertion} & (g,p,\iota)\\
      & \mid & \assume b & \textit{Assumption} & (g,p,\iota) \\
      & \mid & t.\apply() & \textit{Table Application} & (p)\\
      & \mid & c;c & \textit{Sequential Composition} & (a,g,p,\iota) \\
    & \mid & c \choice c & \textit{Nondeterministic Choice} & (g,p,\iota)\\
    & \mid & \choose \rho t c & \textit{Selection} & (\iota) \\
    \multicolumn{3}{l}{
      \begin{array}{ll}
        x \in \Var & h \in \Hdr\\
        n,w,i \in \mathbb{N} & f \in \Field
    \end{array}} & \textit{Sets }\\
  \end{array}
\]
\caption{Grammar of Commands $c \in \Cmd$. Legal action indicated by $a \in
  \Action$, legal instrumented command variants indicated by $\iota \in \Instr$,
  dataplane programs indicated by $p \in \Prog$, and table-free commands
  indicated by $g \in \GCL$.}
\label{fig:grammar}
\end{figure}

@@

We express a collection of invariants on commands as subsets of the set $\Cmd$.
It is convenient to describe these as subgrammars. First, action-commands $a \in
\Action$ contain only non-symbolic assignment and sequence. If an action
command has variables, indicated by the starred variant $a*$, it is called
\emph{open}, if it has none it is called \emph{closed}.

Second, $p \in \Prog \subset \Cmd$ represents the expected set of programs extracted from a
P4 interface, most notably, it contains tables $t$ and no symbolic selection
statements or symbolic expersions.

Conversely $\iota \in \Instr \subset \Cmd$ represents a program instrumented
with symbolic variables. A program $\iota$ cannot apply tables using
$t.\apply()$, but instead uses a havoc statement $\choose \rho t c$, to
introduce symbolic variables $\rho$ in $c$.

Finally $g \in \GCL \subset \Prog \subset \Cmd$ represents the core algebraic
flavor of GCL without havoc, symbolic tables, or table application.

Define $\tables(c) \subset \Table$ to be the set of all tables that occur in $c$.

@@

\begin{figure}
  \[\begin{array}{r>{\triangleq}cl}
  \multicolumn{3}{l}{(\cdot)[\sigma] : \Cmd \to \Cmd} \\
  \multicolumn{3}{l}{(\cdot)[\sigma] : \Prog \to \GCL} \\
  (h.f := e)[\sigma] && h.f := e   \\
  (\assume b)[\sigma] && \assume b \\
  (\assert b)[\sigma] && \assert b \\
  (c_1;c_2)[\sigma] && c_1[\sigma];c_2[\sigma] \\
  (c_1 \choice c_2)[\sigma] && c_1[\sigma] \choice c_2[\sigma] \\
  (\choose \rho t c)[\sigma] && \textrm{\textcolor{red}{unnecessary, but similar to t.apply}}\\
  (t.\apply())[\sigma] &&
  \begin{array}[t]{l}
    \{\\
    \quad \assume {r_1.\matches = t.\keys};\\
    \quad t.\action[r_1.\action](r_1.\data)\\
    \}~\choice\\
    \multicolumn{1}{c}{\vdots}\\
    \choice~\{\\
    \quad \assume {r_n.\matches = t.\keys};\\
    \quad t.\action[r_n.\action](r_n.\data)\\
    \}\\
    \textrm{where}~[r_1;\ldots;r_n] = \sigma(t.\id)
  \end{array}
  \end{array}
  \]
  \caption{Table Elimination}
  \label{fig:table-elim}
\end{figure}

@@

\begin{figure}[htp]
  \[\begin{array}{r >{\triangleq}cl}
  \multicolumn{3}{l}{(\cdot)[?] : \Cmd \to \Cmd} \\
  \multicolumn{3}{l}{(\cdot)[?] : \Prog \to \Instr} \\
  (h.f := e)[?] && h.f := e \\
  (\assume b)[?] && \assume b \\
  (\assert b)[?] && \assert b\\
  (c_1;c_2)[?] && c_1[?];c_2[?]\\
  (c_1 \choice c_2)[?] && c_1[?] \choice c_2[?]\\
  (\choose \rho t c)[?] &&
  \textrm{\textcolor{red}{unnecessary}} \\
  (t.\apply())[?] &&
  \begin{array}[t]{l} \displaystyle
    \{ \\
    \quad \choose \rho t \;\{ \\
    \qquad \assume \rho.\matches = t.\keys; \\
    \qquad \displaystyle \bigchoice_{i=0}^{|t.\action| -1}
    \left\{\begin{array}{l}
    \assume \rho.\action = i;\\
    t.\action[i](\rho.\data)
    \end{array}\right\}\\
    \quad \} \\
    \} \choice \{ \\
    \quad \assume \forall \rho \in t. \rho.\matches \neq t.\keys; \\
    \quad \SKIP \\
    \quad \}\\
    \}
  \end{array}
  \end{array}
  \]
  \caption{Symbolic Table instrumentation}
  \label{fig:table-instrument}
\end{figure}

\begin{lemma}[Models Invariant over Actions]
  \label{lem:models-invariant-over-actions}
  For a closed action command $c$, $\mu,\mu'
  \in \Model$, $\sigma \in \Value \to
  \Row^*$, $\pkt \in \Pkt$, there exists some $\pkt'$
  \[\denote{c}^\sigma\,\mu\,\pkt = \{\mu,\pkt'\}\]
  \end{lemma}

\begin{proof}
  Proceed by induction over $c$.
  \begin{enumerate}[align=left]
  \item[$(c = h.f = e)$] Immediate.
  \item[$(c = c_1;c_2)$]
    By IH($c_1$) there exists some $\pkt''$ such that
    $\denote{c_1}^\sigma\,\mu\,\pkt = \{(\mu,\pkt'')\}$.
    By IH($c_2$) there exists some $\pkt'$ such that
    $\denote{c_2}^\sigma,\mu\,pkt'' = \{(\mu, \pkt')\}$.
    The result follows by definition of $\denote{-}$.
    \end{enumerate}
\end{proof}

\begin{theorem}[Instrumentation Equivalence]
  \label{thm:instr-equiv}
  For every $c \in \Cmd$, well-formed state $\sigma : \Value \to \Row^*$, model
  $\mu \in \Model$, and packet $\pkt \in \Pkt$, $\denote{c}^\sigma~\mu~\pkt =
  \denote{c[?]}^\sigma~\mu~\pkt$.
\end{theorem}

@@

      By Lemma~\ref{lem:models-invariant-over-actions}, there's some $\pkt'$ such that
      \[\{(\mu\{\rho \mapsto r_i\}, \pkt')\} =
      \denote{c[r_i.\data/\vec x]}^\sigma~(\mu\{\rho \mapsto r_i\})~\pkt
      \]

      which means that the result of the first branch is
      \[\{(\mu, \pkt')\} \]

      Since $\rho$ doesn't occur in $c$, then there's some $\mu'$ such that also
      \[\{\mu', \pkt'\} = \denote{c[r_i.\data/\vec x]}^\sigma~\mu~\pkt \]

      and $\mu = \mu'$ by Lemma~\ref{lem:models-invariant-over-actions}.

@@

    \item[\textsc{Miss}] In this case, there is no $r_i$ such that $\pkt,\mu
      \models_\sigma r_i.\matches = t.\keys$, which means that the LHS of the
      toplevel choice operator should evaluate to $\{\}$ \todo{compute this},
      and the RHS evaluates to $\{\mu,\pkt\}$\todo{compute this}, which is
      precisely the result we get when executing
      $\denote{t.apply()}^\sigma~\mu~\pkt$.

@@

\section{The Research Problem}

Now we can formally define the research problems that we want to solve. We have
some ideal goals, that are probably not solvable in general.

\begin{quote}
  \textbf{Conservative Problem.} Given a command $c \in \Cmd$, compute the weakest test
  $b$ such that if a state $\sigma : \Value \to \Row^*$ the following
  judgement is defined and satisfied \[\emptyset, \emptyset \models_\sigma b,\] then \[\forall \pkt \in \Pkt.~
  \denote{c}^\sigma \pkt \neq \error \]
\end{quote}

However, this $b$ may be $\FALSE$ for many realistic programs $c$, so we may
prefer to solve the following problem:

\begin{quote}
  \textbf{Flexible Problem.} Given a command $c \in \Cmd$, compute a test
  $b_{\mathit{cp}}$, and a test $b_{\mathit{pkt}}$ such that
  for every packet $\pkt \in \Pkt$ such that
  \[\emptyset, \emptyset \models_\emptyset b_{\mathit{pkt}},\]
  and for every state $\sigma : \Value \to \Row^*$ such that
  \[\pkt, \emptyset \models_\sigma b_{\mathit{cp}},\]
  then \[\denote{c}^\sigma ~\emptyset~\pkt \neq \error \]
\end{quote}

\section{Reduction to QE}

We can observe that the \textbf{Conservative Problem} is simply solvable. We
sketch the argument here.

\begin{conjecture}[Quantifer Elimination]
  The \textbf{\textrm{Conservative Problem}} is reducable to Quantifer Elimination.
\end{conjecture}

\begin{proof}[Proof Sketch.]
  Given a command $c \in \Cmd$. Compute the instrumented command $c[?]$. We know
  it is equivalent to $c$ by Theorem~\ref{thm:instr-equiv}. Then $\forall
  \overrightarrow{h.f}.\WP(c[?],\TRUE)[\overrightarrow{h.f}]$,\todo{how to write this?} is such a
  formula $b$ by Dijkstra '75. To make $b$ a controller formula, then we simply
  need to remove the dataplane variables via a QE procedure.
\end{proof}

This doesn't work in general. Standard QE procedures are exponential at best, and
enumerating all eth/ipv4 src/dst addresses already takes an intractable amount
of time.

\section{Can Templates save us?}

One option is to use a kind of constrained template-based synthesis, where we
generate a catalogue of temaplates based on syntactic features of the target table.

For example, given a single table like the following:

\[t = \{\keys: [h.f]; \actions: [\lambda x.~~h.g:= x]\}\]

and the simple program \[t.\apply()\]

We may be able to generate a template that looks like

\[\begin{array}{>{\displaystyle}c}
  {[\exists \rho \in t.~ \rho.\keys[0]~\square_{\{=,\neq\}}~?_{v_1}\wedge \rho.\data[0]~\square_{\{=,\neq\}}~?_{v_2} ]?} \\
  \wedge\\
  {[\forall \rho \in t.~\rho.\keys[0]~\square_{\{=,\neq\}}~?_{v_3} \Rightarrow \rho.\data[0]~\square_{\{=,\neq\}}~?_{v_4}]?}
\end{array}\]

where $[b]?$ indicates a hole to either use template $b$ or replace it with
$\TRUE$, $\square_S$ indicates a template for an operation, where $S$ is the set
of allowed operations, and $?_{v_i}$ indicates a hole that should evaluate to an element
of $\Value$.

Now we may wonder: \emph{can we procedurally generate templates in a (nearly) complete way}?

To do so, we need to consider the following dimensions:
\begin{enumerate}[align=left]
\item[\textbf{Keys}.] Should generalize by changing the LHS comparisons to $
  \bigwedge_i[\rho.\keys[i]~\square_{\{=,\neq\}}~?_{v_i}]?$
\item[\textbf{Action Data}.] Should generalize by changing the RHS comparisons to $\bigwedge_i[\rho.\data[i]~\square_{\{=,\neq\}}~?_{v_i}]?$
\item[\textbf{Actions}.] We need to generalize the template across multiple actions. This should be as simple as adding $\rho.\action = i$ on the RHS of each implication above.
\item[\textbf{Sequence}.] I'm not sure how to generalize sequentially composed tables.
  Perhaps a larger catalogue of templates conditioned on structural information:
  if we notice that $t_1$ has a dataflow path to the keys of $t_2$, we should
  generate a join condition template, otherwise we may be able to simply ``and'' the templates.

\item[\textbf{Nondeterminism} ($c_1 \choice c_2$).]. To solve the \textbf{Conservative
  Problem} we simply and the templates together, but to solve the
  \textbf{Flexible Problem} we may be able to constrain the problem (by
  constraining the packets) to either $c_1$ or $c_2$. This would require
  computing a $b_{\textit{pkt}}$.
\end{enumerate}
